\chapter{Efemerideak.}

\section{Sarrera.}

\section{Efemerideak.}

Hiru dira planetak efemerideak,

\begin{enumerate}
\item Jet Propulsion Laboratory, DE (Development Ephemerides).

      Integrazio tartea: $1550-2650$.

      Zenbkizko metodoa: "DIVA" (Krogh,1997).A variable order Adams method.
      
      Doitasuna: "QIVA" a quaruple precision of "DIVA" : the equations of motion , the newtonian part is computed in quadruple precision; all of the rest are computed in double precision.

\item Paris Observatory, INPOP (Intégrateur Númerique Planétaire de l'Observatoire de Paris).
      
      Integrazio tartea: .
      
	  Zenbakizko metodoa: The integrator is an Adams-Cowell method with fixed step-size.
	  
	  Doitasuna:  the programming is done in C language, thus allowing to use the extended precision (80 bits). 
	  Integrating in quadruple precision would of course reduce the round off error in a very large amount, but the
	  CPU time is about 15 time larger than for double precision arithmetic (or extended arithmetic) on our machine
	  (Itanium II with Intel C++ compiler). Nevertheless, it was possible to obtain an additional order of magnitude im-
	  provement by using a single addition in simulated quadruple precision in the corrector step with a very small over-
	  head.
	  
	  Hardware: Intel Itanium II processors.
	  
\item St. Petersburg, EPM (Ephemerides Planets-Moon).
      
      Integrazio tartea: .
      
      Zenbakizko metodoa: Everhart. Implicit RK method (Gauss-Radau).
      (An efficient integrator that uses Gauss-Radau Spacings)
      
      Doitasuna: double precision. The change of ERA system integrator (19 decimal digits instead of 15 ones) with the aim to reduce the round-off error. (Extended precision).
      
\end{enumerate}

\paragraph{Influence of the methods of constructing ephemerides...}
It is obvious that such ephemerides in themselves must have 128 bits, that is, comprise the coefficients calculated with quadruple precision.


\section{Long Integrators.}

